### 2024-10-09 Code/MSB1015-Project.Rmd
Update and rename MSB1015-Project.Rmd to Code/MSB1015-Project.Rmd
Moved code to the folder

---
title: "MSB1015 Project"
author: "Carlo Alberto Zani"
date: "2024-09-08"
output: html_document
---

#0. Setup
```{r setup, include=FALSE}
library(rstudioapi)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(GGally)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()

data <- read_csv("METABRIC_RNA_Mutation.csv")
```

# Data Cleaning
```{r cleaning, echo=FALSE}
# # Check for duplicates
# any(duplicated(data))
# 
# # Check for missing values
# missing_counts <- data %>%
#   summarise(across(everything(), ~ sum(is.na(.)))) %>%
#   pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
#   filter(missing_count > 10)
# 
# # Plot where those missing values were found
# ggplot(missing_counts, aes(x = reorder(column, missing_count), y = missing_count)) +
#   geom_bar(stat = "identity", fill = "skyblue1") +
#   coord_flip() +
#   labs(x = "Column", y = "Missing Count", title = "Columns with More Than 10 Missing Values")
# 
# data$tumor_stage <- as.character(data$tumor_stage)
# 
# data$neoplasm_histologic_grade <- as.character(data$neoplasm_histologic_grade)
# 
# # Apply the modifications
# data_filled <- data %>%
#   
#   # 1. Replace missing values with "Unknown" for all categorical columns
#   mutate(oncotree_code = replace_na(oncotree_code, "Unknown"),
#          cancer_type_detailed = replace_na(cancer_type_detailed, "Unknown" )) %>%
#   
#   mutate(across(where(is.character), ~ factor(replace_na(., "Unknown")))) %>%
#   
#   # 3. Replace missing values with the median for "tumor_size"
#   mutate(tumor_size = ifelse(is.na(tumor_size), median(tumor_size, na.rm = TRUE), tumor_size)) %>%
#   
#   # 4. Replace missing values with 0 for columns ending with "_mut"
#   mutate(across(ends_with("_mut"), ~ replace_na(., 0))) %>%
#   
#   # 5. Convert the columns ending with "_mut" to a logical vector
#   mutate(across(ends_with("_mut"), ~ ifelse(. != "0", 1, 0))) %>%
#   
#   # 6.  
#   rowwise() %>%
#   mutate(mutation_count = sum(across(ends_with("_mut")) == 1))
# 
# # Check for missing values after cleaning
# sum(is.na(data_filled))
# 
# # Select only the transcription data
# pca_gx <- data_filled %>%
#   select(brca1:ugt2b7)
# 
# # Perform PCA to check the columns with the highest amount of missing values
# pca <- prcomp(pca_gx)
# 
# # Convert PCA results to data frame
# pca_data <- as.data.frame(pca$x[, 1:3])
# pca_data$tumor_stage <- as.factor(data_filled$tumor_stage)
# 
# # Create pairwise plots using ggpairs
# ggpairs(pca_data, mapping = aes(color = tumor_stage), 
#         upper = list(continuous = wrap("points")),
#         lower = list(continuous = wrap("points")),
#         diag = list(continuous = wrap("barDiag"))) +
#   theme_minimal()
```

# Imputation
```{r impute, echo=FALSE}

# # Load necessary libraries
# # library(mice)
# 
# # Get rid of unwanted characters from the column names
# colnames(data_filled) <- make.names(colnames(data_filled), unique = TRUE)
# 
# # Convert all character variables to factors
# data_filled <- data_filled %>%
#   mutate(across(where(is.character), as.factor))
# 
# # Wrap 'tumor_stage' as a factor
# data_filled$tumor_stage <- as.factor(data_filled$tumor_stage)
# 
# data_filled$neoplasm_histologic_grade <- as.factor(data_filled$neoplasm_histologic_grade)
# 
# # Identify categorical variables
# fac_vars <- names(data_filled)[sapply(data_filled, is.factor)]
# 
# # Specify methods for all variables, setting CART for categorical variables
# meth <- make.method(data_filled)
# meth[fac_vars] <- "cart" # Use CART for all categorical variables
# 
# # Perform the imputation using mice
# imputation_result <- mice(data_filled, method = meth, m = 5)
# 
# imputed_data <- complete(imputation_result)
# 
# write.csv(imputed_data, file = "imputed_data.csv", row.names = FALSE)
```

# Plots for imputation
```{r Plots after imputation}
# # Convert 'tumor_stage' to a factor in both datasets
# data$tumor_stage <- as.factor(data$tumor_stage)
# imputed_data$tumor_stage <- as.factor(imputed_data$tumor_stage)
# 
# # Create a combined dataset for plotting
# plot_data <- bind_rows(
#   data %>% select(tumor_stage) %>% mutate(Source = 'Original'),
#   imputed_data %>% select(tumor_stage) %>% mutate(Source = 'Imputed')
# )
# 
# # Plot both distributions
# ggplot(data, aes(x = tumor_stage)) +
#   geom_bar(position = "dodge") +
#   labs(title = "Distribution of Tumor Stage: Original vs Imputed",
#        x = "Tumor Stage",
#        y = "Count") +
#   theme_minimal()
```

# 1. Filtering
```{r}
# Check for outliers
imputed_data <- read_csv("imputed_data.csv")

library(isotree)

x<-isolation.forest(imputed_data, ndim = 1, sample_size = 256, max_depth = 8, ntrees = 100, missing_action = "fail", categ_split_type="single_categ", scoring_metric="density")

data_filtered <- imputed_data
data_filtered$pred <- predict(x, imputed_data)

# Create a data frame for plotting
scores_df <- data.frame(
  OutlierScore = data_filtered$pred
)

# Plot the distribution of outlier scores
ggplot(scores_df, aes(x = OutlierScore)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(x = "Outlier Score", y = "Frequency", title = "Distribution of Outlier Scores") +
  theme_minimal()

# Calculate mean and standard deviation
mean_score <- mean(data_filtered$pred, na.rm = TRUE)
sd_score <- sd(data_filtered$pred, na.rm = TRUE)

# Define thresholds for filtering
lower_threshold <- mean_score * 0.95 # might do 5%
upper_threshold <- mean_score * 1.05 # only low visualise with PCA

# Filter out rows where the score is outside of 3 standard deviations from the mean
data_filtered <- data_filtered %>%
  filter(pred >= lower_threshold) %>%
  filter(pred <= upper_threshold) %>%
  select(-pred)

# Check for inconsistencies
result <- lapply(data_filtered, function(i) {
  if (is.factor(i)) {
    unique(i)
  }
})
result <- Filter(Negate(is.null), result)
# Print the result
print(result)
```

# 2. The random forest
```{r random forest, echo=FALSE}
# Load required packages
library(randomForest)
library(caTools)
library(caret)
library(nestedcv)
library(mlbench)
library(pbapply)
library(ROCR)
library(PRROC)
library(pROC)

categorise_prognosis <- function(months) {
  case_when(
    months <= 60 + (60 * 0.25) ~ "will_die_within_5_years",
    months > 60 + (60 * 0.25) ~ "will_die_within_10_years",
  )
}

# Process the data
categorical_data <- data_filtered %>%
  select(age_at_diagnosis:death_from_cancer) %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = categorise_prognosis(overall_survival_months)
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months)

genetic_data <- data_filtered %>%
  select(brca1:stmn2_mut, overall_survival_months, death_from_cancer, overall_survival) %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = categorise_prognosis(overall_survival_months)
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months)

full_data <- data_filtered %>%
  filter(death_from_cancer != "Died of Other Causes") %>%
  filter(overall_survival == 0) %>%
  mutate(
    prognosis = categorise_prognosis(overall_survival_months)
  ) %>%
  select(-death_from_cancer, -overall_survival, -overall_survival_months)

# Split the data into training and testing sets
set.seed(111, "L'Ecuyer-CMRG")

trainIndex <- createDataPartition(categorical_data$prognosis, p = .7, 
                                  list = FALSE, 
                                  times = 1)

train <- categorical_data[ trainIndex,]
test  <- categorical_data[-trainIndex,]

# Ensure "prognosis" is a factor
train$prognosis <- as.factor(train$prognosis)
test$prognosis <- as.factor(test$prognosis)

# Convert train and test features to data frames to avoid tibble row name issues
train_features <- as.data.frame(train %>% select(-prognosis))
test_features <- as.data.frame(test %>% select(-prognosis))
y_train <- train$prognosis

train_sample <- randomsample(y_train, train_features)
y_train <- train_sample$y
x_train <- train_sample$x

# Define the resampling strategy for outer cross-validation
inner_control <- trainControl(method = "cv", number = 5, search = "grid", summaryFunction = twoClassSummary, classProbs = TRUE)

tuneGrid <- expand.grid(mtry = 2:8)
ntrees <- c(1000, 2000) 
nodesize <- c(1, 5)

# List for the combinations
list <- 4

# Define the grid for hyperparameter tuning
params <- expand.grid(ntrees = ntrees,
                      nodesize = nodesize)

store_models <- vector("list", nrow(params))

# Perform nested cross-validation with Random Forest
for(i in 1:nrow(params)) {
  nodesize <- params[i,2]
  ntree <- params[i,1]

  # Train the model with the specified parameters
  nested_result <- nestcv.train(
    x = x_train,
    y = y_train,
    method = "rf",
    savePredictions = "final",
    trControl = inner_control,  # Use the inner cross-validation control
    tuneGrid = tuneGrid,
    ntree = ntree,
    nodesize = nodesize,
    importance = TRUE,
    cv.cores = 8,
    metric = "ROC",
    filterFUN = rf_filter
  )

  # Assign the nested result to store_models and name it based on parameters
  store_models[[i]] <- nested_result
  names(store_models)[i] <- paste0("ntree_", ntree, "_nodesize_", nodesize)
}

# Initialize an empty list to store the metrics
metrics_list <- list()

# Loop over the indices in the provided list
for (i in list) {
  # Apply the metrics function to each model in store_models
  metrics_list[[names(store_models)[i]]] <- metrics(store_models[[i]])
}

# Combine the list of metrics into a dataframe
metrics_df <- do.call(rbind, metrics_list)
metrics_df <- as.data.frame(metrics_df)

# View the resulting dataframe
print(metrics_df)

# Find the model with the highest AUC
best_model_index <- which.max(metrics_df$AUC)

# Access the best model based on the highest ROC
best_model <- store_models[[best_model_index]]

# Get an overview of the best model
summary(best_model)

# Calculate accuracy on the train set for the best model
# model_predict <- best_model$final_fit$finalModel$predicted
# pred_table_training <- table(model_predict, y_train)
# confusionMatrix(pred_table_training)
pred_table_training <- table(best_model$output$predy, best_model$output$testy)
train_confusion <- confusionMatrix(pred_table_training)
train_confusion

# Call the plotting function for the confusion matrix
source("~/University/MSB/MSB1015/confusionPlot.R")
confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)")

# Get predicted probabilities for the positive class ("will_die_within_10_years")
pred_probs <- predict(best_model, newdata = train, type = "prob")
positive_class_probs <- pred_probs[, "will_die_within_10_years"]

# Get the precision recall values
best_model$prc <- prc(best_model)

# Save precision and recall plots as PNG
png("precision_recall_plots_training.png", width = 1600, height = 800)

# Plot precision and recall
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
plot(best_model$roc, col = "red", main = "ROC Training", las = 1)
legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_model$roc), 3)), bty = 'n')
plot(best_model$prc, col = "red", main = "Precision-recall Training")
legend("topright", legend = paste0("AUC = ", round(best_model$prc$auc, 3)), bty = 'n')

# Reset plotting parameters
par(op)

# Close the PNG device
dev.off()

# Extract the importance of the variables for the best model
importance <- as.data.frame(best_model$final_fit$finalModel$importance) %>%
  mutate(will_die_within_10_years = will_die_within_10_years * 10,
         will_die_within_5_years = will_die_within_5_years * 10,
         column = row.names(best_model$final_fit$finalModel$importance))

# Plot the importance of the variables
ggplot(importance, aes(x = reorder(column, will_die_within_10_years))) +
  geom_bar(aes(y = will_die_within_10_years, fill = "Will die within 10 years"), 
           position = "dodge", stat = "identity") +
  geom_bar(aes(y = will_die_within_5_years, fill = "Will die within 5 years"), 
           position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("Will die within 10 years" = "skyblue1", 
                               "Will die within 5 years" = "tomato")) +
  coord_flip() +
  labs(x = "Feature", y = "Importance Score", title = "Variable Importance", fill = "Prognosis") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Extract the error rates (OOB)
err_rate <- as.data.frame(best_model$final_fit$finalModel$err.rate) %>%
  mutate(ntrees = 1:nrow(best_model$final_fit$finalModel$err.rate))

# Plot OOB
ggplot(err_rate, aes(x = ntrees, y = OOB)) +
  geom_line(color = "skyblue1", size = 1) +
  labs(x = "Number of Trees", y = "OBB Error", title = "Out Of Bags (OOB) error") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

###############
### TESTING ###
###############

# Get predicted probabilities for the positive class ("will_die_within_10_years")
pred_probs <- predict(best_model, newdata = test, type = "prob")
positive_class_probs <- pred_probs[, "will_die_within_10_years"]

# Generate the ROC curve
roc_curve <- roc(test$prognosis, positive_class_probs, levels = levels(test$prognosis), direction = ">")

# Create Precision-Recall curve
pr_curve <- pr.curve(scores.class0 = positive_class_probs, 
                     weights.class0 = as.numeric(test$prognosis == "will_die_within_10_years"), 
                     curve = TRUE)

# Save precision and recall plots as PNG
png("precision_recall_plots_train_test.png", width = 1600, height = 800)

# Plot combined graphs
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
plot(best_model$roc, col = "red", main = "ROC", las = 1)
plot(roc_curve, col = "blue", add = TRUE, main = "ROC", las = 1)
legend("bottomright", 
       legend = c(paste0("AUC (Training) = ", signif(pROC::auc(best_model$roc), 2)), 
                  paste0("AUC (Testing) = ", signif(roc_curve$auc, 2))), 
       col = c("red", "blue"), lty = 1, bty = 'n')

# Plot the Precision-Recall curve for the training data
plot(best_model$prc, col = "red", main = "Precision-Recall Curve", xlab = "Recall", ylab = "Precision")
lines(pr_curve$curve[, 3], pr_curve$curve[, 1], col = "blue")
legend("topright", 
       legend = c(paste0("AUC (Training) = ", round(best_model$prc$auc, 2)), 
                  paste0("AUC (Testing) = ", round(pr_curve$auc.integral, 2))), 
       col = c("red", "blue"), 
       lty = 1, 
       bty = 'n')

# Reset plotting parameters and close PNG
par(op)
dev.off()

# Find the best threshold from the ROC curve (maximizing sensitivity-specificity balance)
best_threshold <- as.numeric(coords(roc_curve, "best", ret = "threshold"))
print(paste("Best threshold based on ROC curve:", best_threshold))

# Apply the best threshold for classification
pred_best_threshold <- ifelse(positive_class_probs > best_threshold,
                              "will_die_within_10_years",
                              "will_die_within_5_years")

# Ensure the predicted classes are factors with the correct levels
pred_best_threshold <- factor(pred_best_threshold, levels = levels(test$prognosis))

# Calculate accuracy on the test set with the best threshold
pred_table <- table(pred_best_threshold, test$prognosis)
dimnames(pred_table) <- list(c("will_die_within_10_years", "will_die_within_5_years"), 
                             c("will_die_within_10_years", "will_die_within_5_years"))
test_confusion <- confusionMatrix(pred_table)
test_confusion

confusionPlot(test_confusion, title = "Confusion Matrix (Testing Set)")

```

### 2024-10-10 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Fixed some library issues, and divided the random forest chunk into multiple sub-chunks for easier use. Additionally, I added code to perform unsupervised random forest and visualised it using a function I made.

@@
 
-data <- read_csv("METABRIC_RNA_Mutation.csv")
+# data <- read_csv("METABRIC_RNA_Mutation.csv")
 ```
@@
 
-# Define thresholds for filtering
-lower_threshold <- mean_score * 0.95 # might do 5%
-upper_threshold <- mean_score * 1.05 # only low visualise with PCA
+# Store the original number of rows before filtering
+original_count <- nrow(data_filtered)
 
-# Filter out rows where the score is outside of 3 standard deviations from the mean
+lower_threshold <- mean_score - 2*sd_score
+
+# Filter out rows where the score is outside of 2 standard deviations from the mean
 data_filtered <- data_filtered %>%
   filter(pred >= lower_threshold) %>%
-  filter(pred <= upper_threshold) %>%
-  select(-pred)
+  select(-pred)  # Remove the 'pred' column
 
-# Check for inconsistencies
-result <- lapply(data_filtered, function(i) {
-  if (is.factor(i)) {
-    unique(i)
-  }
-})
-result <- Filter(Negate(is.null), result)
-# Print the result
-print(result)
+# Store the number of rows after filtering
+filtered_count <- nrow(data_filtered)
+
+# Calculate the number of eliminated rows
+eliminated_count <- original_count - filtered_count
+
+# Print the number of eliminated variables
+cat("Number of eliminated variables:", eliminated_count, "\n")
 ```
@@
 # 2. The random forest
-```{r random forest, echo=FALSE}
+```{r data tranformation for random forest, echo=FALSE}
+
 # Load required packages
@@
 library(pROC)
+library(gridExtra)
 
+# Create a function to create the prognosis variable
 categorise_prognosis <- function(months) {
@@
   mutate(
-    prognosis = categorise_prognosis(overall_survival_months)
+    prognosis = as.factor(categorise_prognosis(overall_survival_months))
   ) %>%
-  select(-death_from_cancer, -overall_survival, -overall_survival_months)
+  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
+  as.data.frame()
+
+source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed
+
+# Call the function and plot
+pca_plots_categorical <- pca_proximity_plot(categorical_data, "prognosis")
+grid.arrange(grobs = pca_plots_categorical, ncol = 2)
 
@@
   mutate(
-    prognosis = categorise_prognosis(overall_survival_months)
+    prognosis = as.factor(categorise_prognosis(overall_survival_months))
   ) %>%
-  select(-death_from_cancer, -overall_survival, -overall_survival_months)
+  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
+  as.data.frame()
 
-full_data <- data_filtered %>%
+# Call the function and plot
+pca_plots_genetic <- pca_proximity_plot(genetic_data, "prognosis")
+grid.arrange(grobs = pca_plots_genetic, ncol = 2)
+
+complete_data <- data_filtered %>%
   filter(death_from_cancer != "Died of Other Causes") %>%
@@
   mutate(
-    prognosis = categorise_prognosis(overall_survival_months)
+    prognosis = as.factor(categorise_prognosis(overall_survival_months))
   ) %>%
-  select(-death_from_cancer, -overall_survival, -overall_survival_months)
+  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
+  as.data.frame()
 
-# Split the data into training and testing sets
+# Call the function and plot
+pca_plots_complete <- pca_proximity_plot(complete_data, "prognosis")
+grid.arrange(grobs = pca_plots_complete, ncol = 2)
+```
+
+# 2.1 The random forest
+```{r random forest, echo=FALSE}
+
+# Set the seed for reproducibility of random results using the L'Ecuyer-CMRG method
 set.seed(111, "L'Ecuyer-CMRG")
 
+# Create a partition to split the data into training and testing sets
+# The target variable for stratification is 'prognosis', with 70% of the data for training
 trainIndex <- createDataPartition(categorical_data$prognosis, p = .7, 
-                                  list = FALSE, 
-                                  times = 1)
+                                  list = FALSE,  # Return indices as a matrix
+                                  times = 1)     # Perform the partitioning once
 
-train <- categorical_data[ trainIndex,]
+# Subset the original data to create the training set using the indices from trainIndex
+train <- categorical_data[trainIndex,]
+
+# Subset the original data to create the testing set using the remaining indices
 test  <- categorical_data[-trainIndex,]
@@
 tuneGrid <- expand.grid(mtry = 2:8)
-ntrees <- c(1000, 2000) 
-nodesize <- c(1, 5)
 
-# List for the combinations
-list <- 4
+# The model with the highest AUC that I found was with 2000 trees and node size of 1. 
+# When creating the model these values can be changed.
+ntrees <- c(2000) 
+nodesize <- c(1)
+
+# List for the combinations: should have all the possible combinations between ntrees and nodesize.
+list <- 1
 
@@
 
+```
+
+# 2.2 Finding the best model
+```{r best model, echo=FALSE}
 # Initialize an empty list to store the metrics
@@
 # Loop over the indices in the provided list
-for (i in list) {
+for (i in seq_along(store_models)) {
   # Apply the metrics function to each model in store_models
@@
 summary(best_model)
+```
 
+# 2.3 ROC and PR curves on the best model
+```{r accuracy and plotting, echo=FALSE}
 # Calculate accuracy on the train set for the best model
-# model_predict <- best_model$final_fit$finalModel$predicted
-# pred_table_training <- table(model_predict, y_train)
-# confusionMatrix(pred_table_training)
 pred_table_training <- table(best_model$output$predy, best_model$output$testy)
@@
 # Call the plotting function for the confusion matrix
-source("~/University/MSB/MSB1015/confusionPlot.R")
+source("~/University/MSB/MSB1015/confusionPlot.R") # Change the directory as needed
 confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)")
@@
 
+importance_sorted <- importance[order(importance$will_die_within_10_years, decreasing = TRUE), ]
+head(importance_sorted)
+
 # Plot the importance of the variables
@@
 ggplot(err_rate, aes(x = ntrees, y = OOB)) +
-  geom_line(color = "skyblue1", size = 1) +
+  geom_line(color = "skyblue1", linewidth = 1) +
   labs(x = "Number of Trees", y = "OBB Error", title = "Out Of Bags (OOB) error") +
@@
   theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
+```
 
-###############
-### TESTING ###
-###############
+# 2.4 Testing
+```{r testing, echo=FALSE}
 
@@
 confusionPlot(test_confusion, title = "Confusion Matrix (Testing Set)")
-
 ```

### 2024-10-10 Code/funPCA.R
Add files via upload
The PCA function for the proximity matrix result of the random forest

@@ -1,0 +1,53 @@
+pca_proximity_plot <- function(data, outcome_var, n_trees = 2000, mtry_value = 2, pc_pairs = 1:4) {
+  # Ensure outcome_var is treated as a column name
+  outcome_data <- data[[outcome_var]]
+  
+  # Build Random Forest model with proximity calculation
+  rf_model <- randomForest(x = data[, !names(data) %in% outcome_var], 
+                           mtry = mtry_value, 
+                           ntree = n_trees, 
+                           proximity = TRUE)
+  
+  # Extract proximity matrix and assign row/column names based on the outcome variable
+  proximity_mat <- rf_model$proximity
+  rownames(proximity_mat) <- outcome_data
+  colnames(proximity_mat) <- outcome_data
+  
+  # Function to perform PCA and plot for specified PCs
+  pca_plot <- function(data, pc_x, pc_y) {
+    pca_res <- prcomp(data, center = TRUE, scale. = TRUE)
+    
+    # Create a data frame for the selected principal components
+    pca_df <- data.frame(
+      PCx = pca_res$x[, pc_x],
+      PCy = pca_res$x[, pc_y],
+      Outcome = outcome_data
+    )
+    
+    # Calculate the explained variance
+    explained_var <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100
+    variance_labels <- paste0("PC", c(pc_x, pc_y), ": ", round(explained_var[c(pc_x, pc_y)], 2), "%")
+    
+    # Plot the PCA
+    ggplot(pca_df, aes(x = PCx, y = PCy, color = Outcome)) +
+      geom_point() +
+      labs(
+        title = paste("PCA Plot: PC", pc_x, "vs PC", pc_y),
+        x = variance_labels[1], y = variance_labels[2]
+      ) +
+      theme_bw(base_size = 14) +
+      theme(legend.position = "right")
+  }
+  
+  # Store and create plots for each PCA pair
+  pca_plots <- list()
+  pc_combinations <- combn(pc_pairs, 2, simplify = FALSE)
+  for (i in seq_along(pc_combinations)) {
+    pc_x <- pc_combinations[[i]][1]
+    pc_y <- pc_combinations[[i]][2]
+    pca_plots[[i]] <- pca_plot(proximity_mat, pc_x, pc_y)
+  }
+  
+  # Return the list of PCA plots
+  return(pca_plots)
+}

### 2024-10-14 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Changes in the styling of the code.
Addition of for loops for the random forest process.
Minor updates to the figures.
Made the code more streamlined for multiple datasets.
Removed unused lines.

@@
 author: "Carlo Alberto Zani"
-date: "2024-09-08"
+date: "2024-10-14"
 output: html_document
@@
 
-#0. Setup
+# 0. Setup
 ```{r setup, include=FALSE}
@@
 
-# data <- read_csv("METABRIC_RNA_Mutation.csv")
+data <- read_csv("METABRIC_RNA_Mutation.csv")
 ```
 
-# Data Cleaning
+# 0.1 Data Cleaning
 ```{r cleaning, echo=FALSE}
-# # Check for duplicates
-# any(duplicated(data))
-# 
-# # Check for missing values
-# missing_counts <- data %>%
-#   summarise(across(everything(), ~ sum(is.na(.)))) %>%
-#   pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
-#   filter(missing_count > 10)
-# 
-# # Plot where those missing values were found
-# ggplot(missing_counts, aes(x = reorder(column, missing_count), y = missing_count)) +
-#   geom_bar(stat = "identity", fill = "skyblue1") +
-#   coord_flip() +
-#   labs(x = "Column", y = "Missing Count", title = "Columns with More Than 10 Missing Values")
-# 
-# data$tumor_stage <- as.character(data$tumor_stage)
-# 
-# data$neoplasm_histologic_grade <- as.character(data$neoplasm_histologic_grade)
-# 
-# # Apply the modifications
-# data_filled <- data %>%
-#   
-#   # 1. Replace missing values with "Unknown" for all categorical columns
-#   mutate(oncotree_code = replace_na(oncotree_code, "Unknown"),
-#          cancer_type_detailed = replace_na(cancer_type_detailed, "Unknown" )) %>%
-#   
-#   mutate(across(where(is.character), ~ factor(replace_na(., "Unknown")))) %>%
-#   
-#   # 3. Replace missing values with the median for "tumor_size"
-#   mutate(tumor_size = ifelse(is.na(tumor_size), median(tumor_size, na.rm = TRUE), tumor_size)) %>%
-#   
-#   # 4. Replace missing values with 0 for columns ending with "_mut"
-#   mutate(across(ends_with("_mut"), ~ replace_na(., 0))) %>%
-#   
-#   # 5. Convert the columns ending with "_mut" to a logical vector
-#   mutate(across(ends_with("_mut"), ~ ifelse(. != "0", 1, 0))) %>%
-#   
-#   # 6.  
-#   rowwise() %>%
-#   mutate(mutation_count = sum(across(ends_with("_mut")) == 1))
-# 
-# # Check for missing values after cleaning
-# sum(is.na(data_filled))
-# 
-# # Select only the transcription data
-# pca_gx <- data_filled %>%
-#   select(brca1:ugt2b7)
-# 
-# # Perform PCA to check the columns with the highest amount of missing values
-# pca <- prcomp(pca_gx)
-# 
-# # Convert PCA results to data frame
-# pca_data <- as.data.frame(pca$x[, 1:3])
-# pca_data$tumor_stage <- as.factor(data_filled$tumor_stage)
-# 
-# # Create pairwise plots using ggpairs
-# ggpairs(pca_data, mapping = aes(color = tumor_stage), 
-#         upper = list(continuous = wrap("points")),
-#         lower = list(continuous = wrap("points")),
-#         diag = list(continuous = wrap("barDiag"))) +
-#   theme_minimal()
+# Check for duplicates
+any(duplicated(data))
+
+# Check for missing values
+missing_counts <- data %>%
+  summarise(across(everything(), ~ sum(is.na(.)))) %>%
+  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
+  filter(missing_count > 10)
+
+# Plot where those missing values were found
+ggplot(missing_counts, aes(x = reorder(column, missing_count), y = missing_count)) +
+  geom_bar(stat = "identity", fill = "skyblue1") +
+  coord_flip() +
+  labs(x = "Column", y = "Missing Count", title = "Columns with More Than 10 Missing Values")
+
+data$tumor_stage <- as.factor(data$tumor_stage)
+
+data$neoplasm_histologic_grade <- as.factor(data$neoplasm_histologic_grade)
+
+# Apply the modifications
+data_filled <- data %>%
+  # 1. Replace missing values with "Unknown" for all categorical columns
+  mutate(
+    oncotree_code = replace_na(oncotree_code, "Unknown"),
+    cancer_type_detailed = replace_na(cancer_type_detailed, "Unknown")
+  ) %>%
+  # 3. Replace missing values with the median for "tumor_size"
+  mutate(tumor_size = ifelse(is.na(tumor_size), median(tumor_size, na.rm = TRUE), tumor_size)) %>%
+  # 4. Replace missing values with 0 for columns ending with "_mut"
+  mutate(across(ends_with("_mut"), ~ replace_na(., 0))) %>%
+  # 5. Convert the columns ending with "_mut" to a logical vector
+  mutate(across(ends_with("_mut"), ~ ifelse(. != "0", 1, 0))) %>%
+  # 6.
+  rowwise() %>%
+  mutate(mutation_count = sum(across(ends_with("_mut")) == 1))
+
+# Check for missing values after cleaning
+sum(is.na(data_filled))
+
+# Select only the transcription data
+pca_gx <- data_filled %>%
+  select(brca1:ugt2b7)
+
+# Perform PCA to check the columns with the highest amount of missing values
+pca <- prcomp(pca_gx)
+
+# Convert PCA results to data frame
+pca_data <- as.data.frame(pca$x[, 1:3])
+pca_data$tumor_stage <- as.factor(data_filled$tumor_stage)
+
+# Create pairwise plots using ggpairs
+ggpairs(pca_data,
+  mapping = aes(color = tumor_stage),
+  upper = list(continuous = wrap("points")),
+  lower = list(continuous = wrap("points")),
+  diag = list(continuous = wrap("barDiag"))
+) +
+  theme_minimal()
 ```
 
-# Imputation
+# 0.2 Imputation (this step will take some time)
 ```{r impute, echo=FALSE}
+# Load necessary libraries
+library(mice)
 
-# # Load necessary libraries
-# # library(mice)
-# 
-# # Get rid of unwanted characters from the column names
-# colnames(data_filled) <- make.names(colnames(data_filled), unique = TRUE)
-# 
-# # Convert all character variables to factors
-# data_filled <- data_filled %>%
-#   mutate(across(where(is.character), as.factor))
-# 
-# # Wrap 'tumor_stage' as a factor
-# data_filled$tumor_stage <- as.factor(data_filled$tumor_stage)
-# 
-# data_filled$neoplasm_histologic_grade <- as.factor(data_filled$neoplasm_histologic_grade)
-# 
-# # Identify categorical variables
-# fac_vars <- names(data_filled)[sapply(data_filled, is.factor)]
-# 
-# # Specify methods for all variables, setting CART for categorical variables
-# meth <- make.method(data_filled)
-# meth[fac_vars] <- "cart" # Use CART for all categorical variables
-# 
-# # Perform the imputation using mice
-# imputation_result <- mice(data_filled, method = meth, m = 5)
-# 
-# imputed_data <- complete(imputation_result)
-# 
-# write.csv(imputed_data, file = "imputed_data.csv", row.names = FALSE)
+# Get rid of unwanted characters from the column names
+colnames(data_filled) <- make.names(colnames(data_filled), unique = TRUE)
+
+# Convert all character variables to factors
+data_filled <- data_filled %>%
+  mutate(across(where(is.character), as.factor))
+
+# Identify categorical variables
+fac_vars <- names(data_filled)[sapply(data_filled, is.factor)]
+
+# Specify methods for all variables, setting CART for categorical variables
+meth <- make.method(data_filled)
+meth[fac_vars] <- "cart" # Use CART for all categorical variables
+
+set.seed(111)
+
+# Perform the imputation using the 'mice' package
+imputation_result <- mice(data_filled,
+  method = "cart", # 'meth' contains the method for categorical imputation (CART)
+  m = 5 # 5 is the commonly chosen value for the number of multiple imputations to perform
+)
+
+imputed_data <- complete(imputation_result)
+
+write.csv(imputed_data, file = "imputed_data.csv", row.names = FALSE)
 ```
 
-# Plots for imputation
-```{r Plots after imputation}
-# # Convert 'tumor_stage' to a factor in both datasets
-# data$tumor_stage <- as.factor(data$tumor_stage)
-# imputed_data$tumor_stage <- as.factor(imputed_data$tumor_stage)
-# 
-# # Create a combined dataset for plotting
-# plot_data <- bind_rows(
-#   data %>% select(tumor_stage) %>% mutate(Source = 'Original'),
-#   imputed_data %>% select(tumor_stage) %>% mutate(Source = 'Imputed')
-# )
-# 
-# # Plot both distributions
-# ggplot(data, aes(x = tumor_stage)) +
-#   geom_bar(position = "dodge") +
-#   labs(title = "Distribution of Tumor Stage: Original vs Imputed",
-#        x = "Tumor Stage",
-#        y = "Count") +
-#   theme_minimal()
+# 0.3 Plots for imputation
+```{r plots after imputation, echo=FALSE}
+# Load the imputed data from the previous steps
+imputed_data <- read_csv("imputed_data.csv")
+
+# Convert 'tumor_stage' to a factor in both datasets
+data$tumor_stage <- as.factor(data$tumor_stage)
+imputed_data$tumor_stage <- as.factor(imputed_data$tumor_stage)
+
+# Create a combined dataset for plotting
+plot_data <- bind_rows(
+  data %>% select(tumor_stage) %>% mutate(Source = "Original"),
+  imputed_data %>% select(tumor_stage) %>% mutate(Source = "Imputed")
+)
+
+# Plot the imputed vs original
+ggplot(plot_data, aes(x = tumor_stage, fill = Source)) +
+  geom_bar(position = "dodge") +
+  labs(
+    title = "Distribution of missing values for Tumor Stage: Original vs Imputed",
+    x = "Tumor Stage",
+    y = "Count"
+  ) +
+  theme_minimal() +
+  scale_fill_manual(values = c("Original" = "royalblue", "Imputed" = "firebrick1"))
 ```
@@
 # 1. Filtering
-```{r}
-# Check for outliers
-imputed_data <- read_csv("imputed_data.csv")
-
+```{r isotree filtering, echo=FALSE}
+# You might get an error when loading this package but just run it again and it should work
 library(isotree)
 
-x<-isolation.forest(imputed_data, ndim = 1, sample_size = 256, max_depth = 8, ntrees = 100, missing_action = "fail", categ_split_type="single_categ", scoring_metric="density")
+# In the 'isolation.forest' function parameters were kept as recommended (not default).
+isolation_result <- isolation.forest(imputed_data,
+  ndim = 1, # It was recommended for data with many categorical variables like mine
+  sample_size = 256, # This was the recommended value in most sources
+  max_depth = 8, # ceiling(log2(256)) is the recommended value
+  ntrees = 1000, # The recommended is 100 but the more trees the better usually.
+  missing_action = "fail", # The imputed data has no missing values but in case it does the function would stop
+  categ_split_type = "single_categ", # It was recommended for data with many categorical variables like mine
+  scoring_metric = "depth" # The depth score gives the best performance when looking at literature
+)
 
 data_filtered <- imputed_data
-data_filtered$pred <- predict(x, imputed_data)
+data_filtered$pred <- predict(isolation_result, imputed_data)
 
@@
 ggplot(scores_df, aes(x = OutlierScore)) +
-  geom_histogram(bins = 30, fill = "blue", color = "black") +
+  geom_histogram(bins = 30, fill = "royalblue", color = "black") +
   labs(x = "Outlier Score", y = "Frequency", title = "Distribution of Outlier Scores") +
   theme_minimal()
-
-# Calculate mean and standard deviation
-mean_score <- mean(data_filtered$pred, na.rm = TRUE)
-sd_score <- sd(data_filtered$pred, na.rm = TRUE)
 
@@
 
-lower_threshold <- mean_score - 2*sd_score
-
-# Filter out rows where the score is outside of 2 standard deviations from the mean
+# Filter out rows where the score is above 0.5 which is the standard value for outlier detection
 data_filtered <- data_filtered %>%
-  filter(pred >= lower_threshold) %>%
-  select(-pred)  # Remove the 'pred' column
+  filter(pred <= 0.5) %>%
+  select(-pred)
 
@@
 ```{r data tranformation for random forest, echo=FALSE}
-
 # Load required packages
@@
 
-# Create a function to create the prognosis variable
+# Create a function to form the prognosis variable
 categorise_prognosis <- function(months) {
@@
 
-# Process the data
+# Create the categorical dataframe
 categorical_data <- data_filtered %>%
-  select(age_at_diagnosis:death_from_cancer) %>%
+  select(age_at_diagnosis:death_from_cancer) %>% # Extract the categorical variables
+  filter(death_from_cancer != "Died of Other Causes") %>%
+  filter(overall_survival == 0) %>%
+  mutate(
+    prognosis = as.factor(categorise_prognosis(overall_survival_months)) # Call the function to create 'prognosis'
+  ) %>%
+  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
+  as.data.frame()
+
+source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed
+
+# Call the function for the unsupervised random forest PCA
+pca_plots_categorical <- pca_proximity_plot(categorical_data, "prognosis")
+grid.arrange(grobs = pca_plots_categorical, ncol = 2)
+
+# Create the genetic dataframe
+genetic_data <- data_filtered %>%
+  select(brca1:stmn2_mut, overall_survival_months, death_from_cancer, overall_survival) %>% # Extract the genetic variables and those needed to calculate prognosis
   filter(death_from_cancer != "Died of Other Causes") %>%
@@
 
-source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed
-
-# Call the function and plot
-pca_plots_categorical <- pca_proximity_plot(categorical_data, "prognosis")
-grid.arrange(grobs = pca_plots_categorical, ncol = 2)
-
-genetic_data <- data_filtered %>%
-  select(brca1:stmn2_mut, overall_survival_months, death_from_cancer, overall_survival) %>%
-  filter(death_from_cancer != "Died of Other Causes") %>%
-  filter(overall_survival == 0) %>%
-  mutate(
-    prognosis = as.factor(categorise_prognosis(overall_survival_months))
-  ) %>%
-  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
-  as.data.frame()
-
-# Call the function and plot
+# Call the function for the unsupervised random forest PCA (will take a few minutes)
 pca_plots_genetic <- pca_proximity_plot(genetic_data, "prognosis")
@@
 
+# Create the complete dataframe
 complete_data <- data_filtered %>%
@@
 
-# Call the function and plot
+# Call the function for the unsupervised random forest PCA (will take a few minutes)
 pca_plots_complete <- pca_proximity_plot(complete_data, "prognosis")
@@
 
-# 2.1 The random forest
-```{r random forest, echo=FALSE}
-
-# Set the seed for reproducibility of random results using the L'Ecuyer-CMRG method
+# 2.1 The random forest function
+```{r random forest function, echo=FALSE}
+random_forest <- function(data, ntree, nrun) {
+  # Set the seed for reproducibility using the L'Ecuyer-CMRG method
   set.seed(111, "L'Ecuyer-CMRG")
 
+  # Initialize an empty list to store models from repeated runs
+  store_models_repeated <- vector("list", nrun)
+
+  # Loop to perform nested cross-validation nrun times
+  for (repeat_run in 1:nrun) {
     # Create a partition to split the data into training and testing sets
     # The target variable for stratification is 'prognosis', with 70% of the data for training
-trainIndex <- createDataPartition(categorical_data$prognosis, p = .7, 
+    trainIndex <- createDataPartition(data$prognosis,
+      p = .7,
       list = FALSE, # Return indices as a matrix
-                                  times = 1)     # Perform the partitioning once
+      times = 1
+    ) # Perform the partitioning once
 
-# Subset the original data to create the training set using the indices from trainIndex
-train <- categorical_data[trainIndex,]
-
-# Subset the original data to create the testing set using the remaining indices
-test  <- categorical_data[-trainIndex,]
+    # Subset the original data to create training and testing sets
+    train <- data[trainIndex, ]
+    test <- data[-trainIndex, ]
 
@@
 
-# Convert train and test features to data frames to avoid tibble row name issues
+    # Convert train features to data frames
     train_features <- as.data.frame(train %>% select(-prognosis))
-test_features <- as.data.frame(test %>% select(-prognosis))
     y_train <- train$prognosis
 
+    # Random oversampling of the minority class from 'prognosis'
     train_sample <- randomsample(y_train, train_features)
@@
 
-# Define the resampling strategy for outer cross-validation
-inner_control <- trainControl(method = "cv", number = 5, search = "grid", summaryFunction = twoClassSummary, classProbs = TRUE)
+    # Define the resampling strategy for outer cross-validation.
+    inner_control <- trainControl(
+      method = "cv", # Method was kept as cross validation as standard since it performed the best
+      number = 5, # 5 folds were used for the inner folds as it gave the best result
+      search = "grid", # A grid search was used to test the different tuning variables defined later
+      summaryFunction = twoClassSummary, # Used to retrieve the values need for the ROC and PR curves
+      classProbs = TRUE # Used to retrieve the values need for the ROC and PR curves
+    ) # Everything else was kept standard and NULL as it was not needed.
 
+    # Create a tune grid for the 'mtry' variable
     tuneGrid <- expand.grid(mtry = 2:8)
@@
     # The model with the highest AUC that I found was with 2000 trees and node size of 1.
-# When creating the model these values can be changed.
-ntrees <- c(2000) 
     nodesize <- c(1)
+    params <- expand.grid(ntrees = ntree, nodesize = nodesize)
 
-# List for the combinations: should have all the possible combinations between ntrees and nodesize.
-list <- 1
-
-# Define the grid for hyperparameter tuning
-params <- expand.grid(ntrees = ntrees,
-                      nodesize = nodesize)
-
+    # Store models for the current repeat
     store_models <- vector("list", nrow(params))
 
-# Perform nested cross-validation with Random Forest
-for(i in 1:nrow(params)) {
-  nodesize <- params[i,2]
-  ntree <- params[i,1]
+    # Nested cross-validation loop with Random Forest
+    for (i in 1:nrow(params)) {
+      nodesize <- params[i, 2]
+      ntree <- params[i, 1]
 
-  # Train the model with the specified parameters
+      # Train the model with the specified parameters.
       nested_result <- nestcv.train(
@@
         y = y_train,
-    method = "rf",
-    savePredictions = "final",
+        method = "rf", # Random forest from the 'randomForest' function
+        savePredictions = "final", # 'savePredictions' was kept as final so that it would save the final model with the best scores.
         trControl = inner_control, # Use the inner cross-validation control
-    tuneGrid = tuneGrid,
-    ntree = ntree,
-    nodesize = nodesize,
-    importance = TRUE,
-    cv.cores = 8,
-    metric = "ROC",
-    filterFUN = rf_filter
+        tuneGrid = tuneGrid, # The tune grid is used for the tuning
+        ntree = ntree, # One of the tuned variables
+        nodesize = nodesize, # One of the tuned variables
+        importance = TRUE, # The importance of the variables is scored for later visualisation
+        cv.cores = 8, # Here you can set how many cores to use for parallelisation
+        metric = "ROC", # The 'ROC' metric is used to optimise the models as it is a better representation of performance compared to accuracy.
+        filterFUN = rf_filter # Filtering of the variables helps, especially with the genetic and full databases
       )
 
-  # Assign the nested result to store_models and name it based on parameters
+      # Store the result
       store_models[[i]] <- nested_result
@@
 
-```
+    # Assign the models from this run to store_models_repeated
+    store_models_repeated[[repeat_run]] <- store_models
+  }
 
-# 2.2 Finding the best model
-```{r best model, echo=FALSE}
-# Initialize an empty list to store the metrics
+  # Initialize an empty list to store the metrics for all repetitions
+  metrics_list_repeated <- list()
+
+  # Loop over all repetitions and calculate metrics for each model
+  for (run in seq_along(store_models_repeated)) {
+    store_models <- store_models_repeated[[run]]
+
+    # Initialize an empty list to store the metrics for this repetition
     metrics_list <- list()
 
-# Loop over the indices in the provided list
+    # Loop over the models in the current repetition
     for (i in seq_along(store_models)) {
-  # Apply the metrics function to each model in store_models
+      # Apply the metrics function to each model
       metrics_list[[names(store_models)[i]]] <- metrics(store_models[[i]])
@@
 
-# View the resulting dataframe
-print(metrics_df)
+    # Store the metrics for this repetition
+    metrics_list_repeated[[run]] <- metrics_df
+  }
 
-# Find the model with the highest AUC
+  # Find the best model across all repetitions
+  best_model_overall <- NULL
+  best_auc <- 0
+
+  for (run in seq_along(metrics_list_repeated)) {
+    metrics_df <- metrics_list_repeated[[run]]
+
+    # Find the model with the highest AUC in this repetition
     best_model_index <- which.max(metrics_df$AUC)
 
-# Access the best model based on the highest ROC
-best_model <- store_models[[best_model_index]]
+    # Check if it's the best model overall
+    if (metrics_df$AUC[best_model_index] > best_auc) {
+      best_auc <- metrics_df$AUC[best_model_index]
+      best_model_overall <- store_models_repeated[[run]][[best_model_index]]
+    }
+  }
 
   # Get an overview of the best model
-summary(best_model)
+  print("Best Model Summary")
+  summary(best_model_overall)
+
+  # Return the best model and the list of models for each repetition
+  return(list(best_model_overall = best_model_overall, store_models_repeated = store_models_repeated, train = train, test = test))
+}
 ```
 
-# 2.3 ROC and PR curves on the best model
+# 2.2 Calling the random forest function
+```{r perform random forest, echo=FALSE}
+# Categorical results
+results_categorical <- random_forest(categorical_data, ntree = 10, nrun = 2)
+
+# Genetic results
+results_genetic <- random_forest(genetic_data, ntree = 10, nrun = 2)
+
+# Complete results
+results_complete <- random_forest(complete_data, ntree = 10, nrun = 2)
+```
+
+# 3.1 Create a function to the scores for all the models
+```{r model performance function, echo=FALSE}
+# Create the function to extract the metrics from the result list
+get_final_metrics <- function(results_data) {
+  # Initialize an empty list to store all metrics from all repetitions
+  all_metrics_list <- list()
+
+  # Loop over all repetitions and calculate metrics for each model
+  for (run in seq_along(results_data[[2]])) {
+    store_models <- results_data[[2]][[run]]
+
+    # Initialize an empty list to store the metrics for this repetition
+    metrics_list <- list()
+
+    # Loop over the models in the current repetition
+    for (i in seq_along(store_models)) {
+      # Apply the metrics function to each model
+      metrics_list[[names(store_models)[i]]] <- metrics(store_models[[i]])
+    }
+
+    # Combine the list of metrics into a dataframe
+    metrics_df <- do.call(rbind, metrics_list)
+    metrics_df <- as.data.frame(metrics_df)
+
+    # Add the run information to the metrics dataframe
+    metrics_df$Run <- run # Adding a column to indicate which repetition the metrics belong to
+
+    # Store the metrics for this repetition
+    all_metrics_list[[run]] <- metrics_df
+  }
+
+  # Combine all metrics from all runs into a single dataframe
+  final_metrics_df <- do.call(rbind, all_metrics_list)
+
+  # Transpose the dataframe
+  t_final_metrics_df <- t(final_metrics_df)
+
+  # Convert the matrix to a data frame and keep only the 1st and 3rd rows (AUC, Balanced Accuracy)
+  t_final_metrics_df <- as.data.frame(t_final_metrics_df[c(1, 3), ])
+
+  # Compute the average across the rows
+  t_final_metrics_df <- t_final_metrics_df %>%
+    rowwise() %>%
+    mutate(average = mean(c_across(everything()), na.rm = TRUE)) %>%
+    ungroup() %>%
+    mutate(Metric = c("AUC", "Balanced Accuracy"))
+
+  # Return the final metrics dataframe
+  return(t_final_metrics_df)
+}
+```
+
+# 3.2 Call the function to find the best model
+```{r create the metric dataframes, echo=FALSE}
+# Categorical metrics
+metrics_categorical <- get_final_metrics(results_categorical)
+print(metrics_categorical)
+
+# Genetic metrics
+metrics_genetic <- get_final_metrics(results_genetic)
+print(metrics_genetic)
+
+# Complete metrics
+metrics_complete <- get_final_metrics(results_complete)
+print(metrics_complete)
+
+# Add class column to each dataframe
+metrics_categorical$class <- "categorical"
+metrics_genetic$class <- "genetic"
+metrics_complete$class <- "complete"
+
+# Combine all dataframes into one
+all_metrics <- bind_rows(metrics_categorical, metrics_genetic, metrics_complete)
+
+# Extract only the rows corresponding to AUC
+auc_metrics <- all_metrics %>%
+  filter(Metric == "AUC") %>%
+  filter(average == max(average))
+
+# Extract and print the class of the best model
+best_model_class <- auc_metrics$class
+
+# Create the name of the dataset corresponding to the best model class
+best_class_name <- paste("results_", best_model_class, sep = "")
+
+# Print the best class name
+print(best_class_name)
+
+# Get the the name of the data with the best model
+best_class_data <- get(best_class_name)
+```
+
+# 4. ROC and PR curves on the best model from the best class
 ```{r accuracy and plotting, echo=FALSE}
 # Calculate accuracy on the train set for the best model
-pred_table_training <- table(best_model$output$predy, best_model$output$testy)
+pred_table_training <- table(best_class_data$best_model_overall$output$predy, best_class_data$best_model_overall$output$testy)
 train_confusion <- confusionMatrix(pred_table_training)
@@
 # Get predicted probabilities for the positive class ("will_die_within_10_years")
-pred_probs <- predict(best_model, newdata = train, type = "prob")
+pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$train, type = "prob")
 positive_class_probs <- pred_probs[, "will_die_within_10_years"]
@@
 # Get the precision recall values
-best_model$prc <- prc(best_model)
+best_class_data$best_model_overall$prc <- prc(best_class_data$best_model_overall)
 
@@
 op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
-plot(best_model$roc, col = "red", main = "ROC Training", las = 1)
-legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_model$roc), 3)), bty = 'n')
-plot(best_model$prc, col = "red", main = "Precision-recall Training")
-legend("topright", legend = paste0("AUC = ", round(best_model$prc$auc, 3)), bty = 'n')
+plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC Training", las = 1)
+legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 3)), bty = "n")
+plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-recall Training")
+legend("topright", legend = paste0("AUC = ", round(best_class_data$best_model_overall$prc$auc, 3)), bty = "n")
 
@@
 # Extract the importance of the variables for the best model
-importance <- as.data.frame(best_model$final_fit$finalModel$importance) %>%
-  mutate(will_die_within_10_years = will_die_within_10_years * 10,
+importance <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$importance) %>%
+  mutate(
+    will_die_within_10_years = will_die_within_10_years * 10,
     will_die_within_5_years = will_die_within_5_years * 10,
-         column = row.names(best_model$final_fit$finalModel$importance))
+    column = row.names(best_class_data$best_model_overall$final_fit$finalModel$importance)
+  )
 
@@
   geom_bar(aes(y = will_die_within_10_years, fill = "Will die within 10 years"),
-           position = "dodge", stat = "identity") +
+    position = "dodge", stat = "identity"
+  ) +
   geom_bar(aes(y = will_die_within_5_years, fill = "Will die within 5 years"),
-           position = "dodge", stat = "identity") +
-  scale_fill_manual(values = c("Will die within 10 years" = "skyblue1", 
-                               "Will die within 5 years" = "tomato")) +
+    position = "dodge", stat = "identity"
+  ) +
+  scale_fill_manual(values = c(
+    "Will die within 10 years" = "skyblue1",
+    "Will die within 5 years" = "tomato"
+  )) +
   coord_flip() +
@@
 # Extract the error rates (OOB)
-err_rate <- as.data.frame(best_model$final_fit$finalModel$err.rate) %>%
-  mutate(ntrees = 1:nrow(best_model$final_fit$finalModel$err.rate))
+err_rate <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$err.rate) %>%
+  mutate(ntrees = 1:nrow(best_class_data$best_model_overall$final_fit$finalModel$err.rate))
 
@@
 
-# 2.4 Testing
+# 5. Testing on the best model
 ```{r testing, echo=FALSE}
-
 # Get predicted probabilities for the positive class ("will_die_within_10_years")
-pred_probs <- predict(best_model, newdata = test, type = "prob")
+pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$test, type = "prob")
 positive_class_probs <- pred_probs[, "will_die_within_10_years"]
@@
 # Generate the ROC curve
-roc_curve <- roc(test$prognosis, positive_class_probs, levels = levels(test$prognosis), direction = ">")
+roc_curve <- roc(best_class_data$test$prognosis, positive_class_probs, levels = levels(best_class_data$test$prognosis), direction = ">")
 
 # Create Precision-Recall curve
-pr_curve <- pr.curve(scores.class0 = positive_class_probs, 
-                     weights.class0 = as.numeric(test$prognosis == "will_die_within_10_years"), 
-                     curve = TRUE)
+pr_curve <- pr.curve(
+  scores.class0 = positive_class_probs,
+  weights.class0 = as.numeric(best_class_data$test$prognosis == "will_die_within_10_years"),
+  curve = TRUE
+)
 
@@
 op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
-plot(best_model$roc, col = "red", main = "ROC", las = 1)
+plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC", las = 1)
 plot(roc_curve, col = "blue", add = TRUE, main = "ROC", las = 1)
 legend("bottomright",
-       legend = c(paste0("AUC (Training) = ", signif(pROC::auc(best_model$roc), 2)), 
-                  paste0("AUC (Testing) = ", signif(roc_curve$auc, 2))), 
-       col = c("red", "blue"), lty = 1, bty = 'n')
+  legend = c(
+    paste0("AUC (Training) = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 2)),
+    paste0("AUC (Testing) = ", signif(roc_curve$auc, 2))
+  ),
+  col = c("red", "blue"), lty = 1, bty = "n"
+)
 
 # Plot the Precision-Recall curve for the training data
-plot(best_model$prc, col = "red", main = "Precision-Recall Curve", xlab = "Recall", ylab = "Precision")
+plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-Recall Curve", xlab = "Recall", ylab = "Precision")
 lines(pr_curve$curve[, 3], pr_curve$curve[, 1], col = "blue")
 legend("topright",
-       legend = c(paste0("AUC (Training) = ", round(best_model$prc$auc, 2)), 
-                  paste0("AUC (Testing) = ", round(pr_curve$auc.integral, 2))), 
+  legend = c(
+    paste0("AUC (Training) = ", round(best_class_data$best_model_overall$prc$auc, 2)),
+    paste0("AUC (Testing) = ", round(pr_curve$auc.integral, 2))
+  ),
   col = c("red", "blue"),
   lty = 1,
-       bty = 'n')
+  bty = "n"
+)
 
@@
   "will_die_within_10_years",
-                              "will_die_within_5_years")
+  "will_die_within_5_years"
+)
 
 # Ensure the predicted classes are factors with the correct levels
-pred_best_threshold <- factor(pred_best_threshold, levels = levels(test$prognosis))
+pred_best_threshold <- factor(pred_best_threshold, levels = levels(best_class_data$test$prognosis))
 
 # Calculate accuracy on the test set with the best threshold
-pred_table <- table(pred_best_threshold, test$prognosis)
-dimnames(pred_table) <- list(c("will_die_within_10_years", "will_die_within_5_years"), 
-                             c("will_die_within_10_years", "will_die_within_5_years"))
+pred_table <- table(pred_best_threshold, best_class_data$test$prognosis)
+dimnames(pred_table) <- list(
+  c("will_die_within_10_years", "will_die_within_5_years"),
+  c("will_die_within_10_years", "will_die_within_5_years")
+)
 test_confusion <- confusionMatrix(pred_table)

### 2024-10-14 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Added an if statement to install the required libraries

@@
 ```{r setup, include=FALSE}
+# List of required packages
+required_packages <- c("rstudioapi", "readr", "dplyr", "ggplot2", "tidyr", 
+                       "stringr", "GGally", "mice", "isotree", "randomForest", 
+                       "caTools", "caret", "nestedcv", "mlbench", "pbapply", 
+                       "ROCR", "PRROC", "pROC", "gridExtra")
+
+# Check and install missing packages
+new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
+if (length(new_packages)) {
+  install.packages(new_packages)
+}
+
+# Load libraries
 library(rstudioapi)

### 2024-10-14 Code/confusionPlot.R
Update confusionPlot.R
Added an if statement to install libraries

@@ -1,3 +1,12 @@
+# List of additional required packages
+additional_packages <- c("caret", "ggplot2", "reshape2", "wesanderson")
+
+# Check and install missing packages
+new_additional_packages <- additional_packages[!(additional_packages %in% installed.packages()[,"Package"])]
+if (length(new_additional_packages)) {
+  install.packages(new_additional_packages)
+}
+
 # Load necessary libraries
 library(caret)
 library(ggplot2)

### 2024-10-14 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Added more comments and scaled for the initial PCA

--- 2024-10-14 Code/MSB1015-Project.Rmd
@@
 ```{r setup, include=FALSE}
+# List of required packages
+required_packages <- c(
+  "rstudioapi", "readr", "dplyr", "ggplot2", "tidyr",
+  "stringr", "GGally", "mice", "isotree", "randomForest",
+  "caTools", "caret", "nestedcv", "mlbench", "pbapply",
+  "ROCR", "PRROC", "pROC", "gridExtra"
+)
+
+# Check and install missing packages
+new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
+if (length(new_packages)) {
+  install.packages(new_packages)
+}
+
+# Load libraries
 library(rstudioapi)
@@
 library(GGally)
+library(mice)
+library(randomForest)
+library(caTools)
+library(caret)
+library(nestedcv)
+library(mlbench)
+library(pbapply)
+library(ROCR)
+library(PRROC)
+library(pROC)
+library(gridExtra)
 
+# Set the working directory to the one where this code is
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
@@
 
+# Load the data
 data <- read_csv("METABRIC_RNA_Mutation.csv")
@@
 missing_counts <- data %>%
-  summarise(across(everything(), ~ sum(is.na(.)))) %>%
-  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
-  filter(missing_count > 10)
+  summarise(across(everything(), ~ sum(is.na(.)))) %>% # Summarise missing values per column
+  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>% # Reshape into long format
+  filter(missing_count > 10) # Filter columns with more than 10 missing values
 
@@
 ggplot(missing_counts, aes(x = reorder(column, missing_count), y = missing_count)) +
-  geom_bar(stat = "identity", fill = "skyblue1") +
+  geom_bar(stat = "identity", fill = "royalblue") +
   coord_flip() +
@@
 
+# Convert specific columns to factors
 data$tumor_stage <- as.factor(data$tumor_stage)
-
 data$neoplasm_histologic_grade <- as.factor(data$neoplasm_histologic_grade)
@@
 data_filled <- data %>%
-  # 1. Replace missing values with "Unknown" for all categorical columns
+  # 1. Replace missing values with "Unknown" for specific categorical columns (not suitable for imputation)
   mutate(
@@
   ) %>%
-  # 3. Replace missing values with the median for "tumor_size"
+  # 2. Replace missing values in "tumor_size" with the median value of the column
   mutate(tumor_size = ifelse(is.na(tumor_size), median(tumor_size, na.rm = TRUE), tumor_size)) %>%
-  # 4. Replace missing values with 0 for columns ending with "_mut"
+  # 3. Replace missing values with 0 for all columns that end with "_mut"
   mutate(across(ends_with("_mut"), ~ replace_na(., 0))) %>%
-  # 5. Convert the columns ending with "_mut" to a logical vector
+  # 4. Convert values in columns ending with "_mut" to a binary vector (values not equal to "0" are converted to 1)
   mutate(across(ends_with("_mut"), ~ ifelse(. != "0", 1, 0))) %>%
-  # 6.
+  # 5. Calculate the "mutation_count" by summing the binary values (1s) in columns that end with "_mut" for each row
   rowwise() %>%
@@
 # Perform PCA to check the columns with the highest amount of missing values
-pca <- prcomp(pca_gx)
+pca <- prcomp(pca_gx, center = TRUE, scale. = TRUE)
 
@@
 ```{r impute, echo=FALSE}
-# Load necessary libraries
-library(mice)
-
-# Get rid of unwanted characters from the column names
+# Get rid of unwanted characters from the column names (the sign '+' from one of the variables)
 colnames(data_filled) <- make.names(colnames(data_filled), unique = TRUE)
@@
   mutate(across(where(is.character), as.factor))
-
-# Identify categorical variables
-fac_vars <- names(data_filled)[sapply(data_filled, is.factor)]
-
-# Specify methods for all variables, setting CART for categorical variables
-meth <- make.method(data_filled)
-meth[fac_vars] <- "cart" # Use CART for all categorical variables
 
@@
 
+# Collect the results
 imputed_data <- complete(imputation_result)
 
+# Save the imputed data in the working directory
 write.csv(imputed_data, file = "imputed_data.csv", row.names = FALSE)
@@
 ```{r data tranformation for random forest, echo=FALSE}
-# Load required packages
-library(randomForest)
-library(caTools)
-library(caret)
-library(nestedcv)
-library(mlbench)
-library(pbapply)
-library(ROCR)
-library(PRROC)
-library(pROC)
-library(gridExtra)
-
-# Create a function to form the prognosis variable
+# Create a function to form the prognosis variable based on overall survival months
 categorise_prognosis <- function(months) {
@@
     months <= 60 + (60 * 0.25) ~ "will_die_within_5_years",
-    months > 60 + (60 * 0.25) ~ "will_die_within_10_years",
+    months > 60 + (60 * 0.25) ~ "will_die_within_10_years"
   )
@@
   select(age_at_diagnosis:death_from_cancer) %>% # Extract the categorical variables
-  filter(death_from_cancer != "Died of Other Causes") %>%
-  filter(overall_survival == 0) %>%
+  filter(death_from_cancer != "Died of Other Causes") %>% # Remove records for non-cancer deaths
+  filter(overall_survival == 0) %>% # Keep only the rows where overall survival is 0
   mutate(
-    prognosis = as.factor(categorise_prognosis(overall_survival_months)) # Call the function to create 'prognosis'
+    prognosis = as.factor(categorise_prognosis(overall_survival_months)) # Apply function to create 'prognosis'
   ) %>%
-  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>%
-  as.data.frame()
+  select(-death_from_cancer, -overall_survival, -overall_survival_months) %>% # Drop unnecessary columns
+  as.data.frame() # Convert to a dataframe
 
+# Load the function for PCA
 source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed
@@
 ```{r random forest function, echo=FALSE}
+# Create the function to perform random forest classification
 random_forest <- function(data, ntree, nrun) {
@@
   t_final_metrics_df <- t_final_metrics_df %>%
-    rowwise() %>%
-    mutate(average = mean(c_across(everything()), na.rm = TRUE)) %>%
+    rowwise() %>% # Perform row-wise operations
+    mutate(average = mean(c_across(everything()), na.rm = TRUE)) %>% # Calculate row-wise average
     ungroup() %>%
-    mutate(Metric = c("AUC", "Balanced Accuracy"))
+    mutate(Metric = c("AUC", "Balanced Accuracy")) # Add a Metric column
 
@@
 # Calculate accuracy on the train set for the best model
-pred_table_training <- table(best_class_data$best_model_overall$output$predy, best_class_data$best_model_overall$output$testy)
-train_confusion <- confusionMatrix(pred_table_training)
+pred_table_training <- table(best_class_data$best_model_overall$output$predy, best_class_data$best_model_overall$output$testy) # Create prediction table
+train_confusion <- confusionMatrix(pred_table_training) # Compute confusion matrix
 train_confusion
@@
 # Call the plotting function for the confusion matrix
-source("~/University/MSB/MSB1015/confusionPlot.R") # Change the directory as needed
-confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)")
+source("~/University/MSB/MSB1015/confusionPlot.R") # Load external plot function (update directory if needed)
+confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)") # Plot confusion matrix
 
 # Get predicted probabilities for the positive class ("will_die_within_10_years")
-pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$train, type = "prob")
-positive_class_probs <- pred_probs[, "will_die_within_10_years"]
+pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$train, type = "prob") # Predict probabilities
+positive_class_probs <- pred_probs[, "will_die_within_10_years"] # Extract probabilities for "will_die_within_10_years"
 
-# Get the precision recall values
-best_class_data$best_model_overall$prc <- prc(best_class_data$best_model_overall)
+# Get the precision-recall values
+best_class_data$best_model_overall$prc <- prc(best_class_data$best_model_overall) # Calculate precision-recall
 
@@
 # Plot precision and recall
-op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
-plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC Training", las = 1)
-legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 3)), bty = "n")
-plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-recall Training")
-legend("topright", legend = paste0("AUC = ", round(best_class_data$best_model_overall$prc$auc, 3)), bty = "n")
+op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1) # Set up the plotting layout
+plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC Training", las = 1) # Plot ROC curve
+legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 3)), bty = "n") # Add AUC legend
+plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-recall Training") # Plot Precision-Recall curve
+legend("topright", legend = paste0("AUC = ", round(best_class_data$best_model_overall$prc$auc, 3)), bty = "n") # Add AUC legend for PRC
 
@@
 # Extract the importance of the variables for the best model
-importance <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$importance) %>%
+importance <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$importance) %>% # Get variable importance
   mutate(
-    will_die_within_10_years = will_die_within_10_years * 10,
+    will_die_within_10_years = will_die_within_10_years * 10, # Scale the importance scores
     will_die_within_5_years = will_die_within_5_years * 10,
-    column = row.names(best_class_data$best_model_overall$final_fit$finalModel$importance)
+    column = row.names(best_class_data$best_model_overall$final_fit$finalModel$importance) # Extract feature names
   )
 
-importance_sorted <- importance[order(importance$will_die_within_10_years, decreasing = TRUE), ]
+importance_sorted <- importance[order(importance$will_die_within_10_years, decreasing = TRUE), ] # Sort by importance
 head(importance_sorted)
@@
   theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
+  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) # Create variable importance plot
 
 # Extract the error rates (OOB)
-err_rate <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$err.rate) %>%
-  mutate(ntrees = 1:nrow(best_class_data$best_model_overall$final_fit$finalModel$err.rate))
+err_rate <- as.data.frame(best_class_data$best_model_overall$final_fit$finalModel$err.rate) %>% # Extract error rates
+  mutate(ntrees = 1:nrow(best_class_data$best_model_overall$final_fit$finalModel$err.rate)) # Add number of trees
 
-# Plot OOB
+# Plot OOB error
 ggplot(err_rate, aes(x = ntrees, y = OOB)) +
   geom_line(color = "skyblue1", linewidth = 1) +
-  labs(x = "Number of Trees", y = "OBB Error", title = "Out Of Bags (OOB) error") +
+  labs(x = "Number of Trees", y = "OOB Error", title = "Out Of Bag (OOB) Error") +
   theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
+  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) # Plot OOB error
 ```
@@
 pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$test, type = "prob")
-positive_class_probs <- pred_probs[, "will_die_within_10_years"]
+positive_class_probs <- pred_probs[, "will_die_within_10_years"] # Select probabilities for "will_die_within_10_years"
 
-# Generate the ROC curve
+# Generate the ROC curve for the test set
 roc_curve <- roc(best_class_data$test$prognosis, positive_class_probs, levels = levels(best_class_data$test$prognosis), direction = ">")
 
-# Create Precision-Recall curve
+# Create Precision-Recall curve for the test set
 pr_curve <- pr.curve(
-  scores.class0 = positive_class_probs,
-  weights.class0 = as.numeric(best_class_data$test$prognosis == "will_die_within_10_years"),
-  curve = TRUE
+  scores.class0 = positive_class_probs, # Probabilities for the positive class
+  weights.class0 = as.numeric(best_class_data$test$prognosis == "will_die_within_10_years"), # Binary vector for positive class
+  curve = TRUE # Return curve for plotting
 )
@@
 # Save precision and recall plots as PNG
-png("precision_recall_plots_train_test.png", width = 1600, height = 800)
+png("precision_recall_plots_train_test.png", width = 1600, height = 800) # Save output to PNG file
 
-# Plot combined graphs
-op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1)
-plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC", las = 1)
-plot(roc_curve, col = "blue", add = TRUE, main = "ROC", las = 1)
-legend("bottomright",
+# Plot combined graphs (ROC and Precision-Recall curves side by side)
+op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 2) + 0.1) # Set layout to 1 row, 2 columns
+
+# Plot ROC curve for both training and testing
+plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC", las = 1) # Training ROC
+plot(roc_curve, col = "blue", add = TRUE, main = "ROC", las = 1) # Testing ROC
+legend("bottomright", # Add legend with AUC values
   legend = c(
@@
 
-# Plot the Precision-Recall curve for the training data
-plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-Recall Curve", xlab = "Recall", ylab = "Precision")
-lines(pr_curve$curve[, 3], pr_curve$curve[, 1], col = "blue")
-legend("topright",
+# Plot Precision-Recall curve for both training and testing
+plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-Recall Curve", xlab = "Recall", ylab = "Precision") # Training PRC
+lines(pr_curve$curve[, 3], pr_curve$curve[, 1], col = "blue") # Testing PRC
+legend("topright", # Add legend with AUC values
   legend = c(
@@
 
-# Reset plotting parameters and close PNG
+# Reset plotting parameters and close PNG file
 par(op)
@@
 
-# Find the best threshold from the ROC curve (maximizing sensitivity-specificity balance)
-best_threshold <- as.numeric(coords(roc_curve, "best", ret = "threshold"))
+# Find the best threshold from the ROC curve
+best_threshold <- as.numeric(coords(roc_curve, "best", ret = "threshold")) # Threshold maximizing sensitivity-specificity
 print(paste("Best threshold based on ROC curve:", best_threshold))
@@
 # Apply the best threshold for classification
-pred_best_threshold <- ifelse(positive_class_probs > best_threshold,
-  "will_die_within_10_years",
-  "will_die_within_5_years"
-)
+pred_best_threshold <- ifelse(positive_class_probs > best_threshold, "will_die_within_10_years", "will_die_within_5_years") # Classify based on threshold
 
-# Ensure the predicted classes are factors with the correct levels
+# Ensure predicted classes have the correct factor levels
 pred_best_threshold <- factor(pred_best_threshold, levels = levels(best_class_data$test$prognosis))
 
-# Calculate accuracy on the test set with the best threshold
-pred_table <- table(pred_best_threshold, best_class_data$test$prognosis)
-dimnames(pred_table) <- list(
+# Calculate accuracy on the test set using the best threshold
+pred_table <- table(pred_best_threshold, best_class_data$test$prognosis) # Create confusion matrix table
+dimnames(pred_table) <- list( # Assign dimension names to match factor levels
   c("will_die_within_10_years", "will_die_within_5_years"),
@@
 )
-test_confusion <- confusionMatrix(pred_table)
+test_confusion <- confusionMatrix(pred_table) # Compute confusion matrix
 test_confusion
 
+# Plot the confusion matrix for the test set
 confusionPlot(test_confusion, title = "Confusion Matrix (Testing Set)")

### 2024-10-17 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd

@@
 plot(best_class_data$best_model_overall$roc, col = "red", main = "ROC Training", las = 1) # Plot ROC curve
-legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 3)), bty = "n") # Add AUC legend
+legend("bottomright", legend = paste0("AUC = ", signif(pROC::auc(best_class_data$best_model_overall$roc), 2)), bty = "n") # Add AUC legend
 plot(best_class_data$best_model_overall$prc, col = "red", main = "Precision-recall Training") # Plot Precision-Recall curve
-legend("topright", legend = paste0("AUC = ", round(best_class_data$best_model_overall$prc$auc, 3)), bty = "n") # Add AUC legend for PRC
+legend("topright", legend = paste0("AUC = ", round(best_class_data$best_model_overall$prc$auc, 2)), bty = "n") # Add AUC legend for PRC
 

### 2024-10-18 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Added a chunk to find the matching train and test splits with the best model

@@
   coord_flip() +
-  labs(x = "Column", y = "Missing Count", title = "Columns with More Than 10 Missing Values")
+  labs(x = "Column", y = "Missing Count", title = "Columns with More Than 10 Missing Values") +
+  theme(
+    panel.background = element_rect(fill = "white", color = "white"),
+    plot.background = element_rect(fill = "white", color = "white"),
+    panel.grid.major = element_line(color = "gray80"),
+    panel.grid.minor = element_line(color = "gray90"),
+    axis.text.y = element_text(size = 14),  
+    axis.text.x = element_text(size = 12),  
+    axis.title = element_text(size = 14),
+    plot.title = element_text(size = 15, face = "bold")
+  )
 
@@
 
+  # Initialize lists to store all train and test splits
+  train_splits <- vector("list", nrun)
+  test_splits <- vector("list", nrun)
+
   # Loop to perform nested cross-validation nrun times
@@
     test <- data[-trainIndex, ]
+
+    # Store the train and test sets for this run
+    train_splits[[repeat_run]] <- train
+    test_splits[[repeat_run]] <- test
 
@@
   # Return the best model and the list of models for each repetition
-  return(list(best_model_overall = best_model_overall, store_models_repeated = store_models_repeated, train = train, test = test))
+  return(list(best_model_overall = best_model_overall, 
+              store_models_repeated = store_models_repeated, 
+              train_splits = train_splits, 
+              test_splits = test_splits
+              ))
 }
@@
 # Categorical results
-results_categorical <- random_forest(categorical_data, ntree = 10, nrun = 2)
+results_categorical <- random_forest(categorical_data, ntree = 1000, nrun = 5)
 
 # Genetic results
-results_genetic <- random_forest(genetic_data, ntree = 10, nrun = 2)
+results_genetic <- random_forest(genetic_data, ntree = 10, nrun = 1)
 
 # Complete results
-results_complete <- random_forest(complete_data, ntree = 10, nrun = 2)
+results_complete <- random_forest(complete_data, ntree = 10, nrun = 1)
 ```
@@
 
+# 3.3 Finding the matching splits for the best model
+```{r finding matching splits}
+# Get the AUC value of the best model
+best_auc <- best_class_data$best_model_overall$roc$auc
+
+# Initialize variables to store the matching run and model
+matching_run <- NULL
+matching_model <- NULL
+matching_index <- NULL
+
+# Loop over all runs in store_models_repeated
+for (run_index in seq_along(best_class_data$store_models_repeated)) {
+  # Get the current run's list of models
+  current_run_models <- best_class_data$store_models_repeated[[run_index]]
+
+  # Loop over all models in the current run and compare their AUC values
+  for (model_index in seq_along(current_run_models)) {
+    model_name <- names(current_run_models)[model_index]
+    model_auc <- current_run_models[[model_name]]$roc$auc
+
+    # Check if the AUC matches the best model's AUC
+    if (model_auc == best_auc) {
+      matching_run <- run_index
+      matching_model <- model_name
+      matching_index <- model_index
+      break
+    }
+  }
+
+  # Break out of the outer loop if a match is found
+  if (!is.null(matching_run)) {
+    break
+  }
+}
+
+# If a match is found, print the run, model name, and retrieve the corresponding train/test splits
+if (!is.null(matching_run) && !is.null(matching_model)) {
+  print(paste("Matching model found in run:", matching_run, "model:", matching_model))
+
+  # Retrieve the corresponding train and test splits
+  matching_train_split <- best_class_data$train_splits[[matching_run]]
+  matching_test_split <- best_class_data$test_splits[[matching_run]]
+
+  # Print or use the splits as needed
+  print("Found Matching Splits")
+
+} else {
+  print("No matching model found")
+}
+
+```
+
 # 4. ROC and PR curves on the best model from the best class
@@
 # Get predicted probabilities for the positive class ("will_die_within_10_years")
-pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$train, type = "prob") # Predict probabilities
+pred_probs <- predict(best_class_data$best_model_overall, newdata = matching_train_split, type = "prob") # Predict probabilities
 positive_class_probs <- pred_probs[, "will_die_within_10_years"] # Extract probabilities for "will_die_within_10_years"
@@
 # Get predicted probabilities for the positive class ("will_die_within_10_years")
-pred_probs <- predict(best_class_data$best_model_overall, newdata = best_class_data$test, type = "prob")
+pred_probs <- predict(best_class_data$best_model_overall, newdata = matching_test_split, type = "prob")
 positive_class_probs <- pred_probs[, "will_die_within_10_years"] # Select probabilities for "will_die_within_10_years"
@@
 # Generate the ROC curve for the test set
-roc_curve <- roc(best_class_data$test$prognosis, positive_class_probs, levels = levels(best_class_data$test$prognosis), direction = ">")
+roc_curve <- roc(matching_test_split$prognosis, positive_class_probs, levels = levels(matching_test_split$prognosis), direction = ">")
 
@@
   scores.class0 = positive_class_probs, # Probabilities for the positive class
-  weights.class0 = as.numeric(best_class_data$test$prognosis == "will_die_within_10_years"), # Binary vector for positive class
+  weights.class0 = as.numeric(matching_test_split$prognosis == "will_die_within_10_years"), # Binary vector for positive class
   curve = TRUE # Return curve for plotting
@@
 # Ensure predicted classes have the correct factor levels
-pred_best_threshold <- factor(pred_best_threshold, levels = levels(best_class_data$test$prognosis))
+pred_best_threshold <- factor(pred_best_threshold, levels = levels(matching_test_split$prognosis))
 
 # Calculate accuracy on the test set using the best threshold
-pred_table <- table(pred_best_threshold, best_class_data$test$prognosis) # Create confusion matrix table
+pred_table <- table(pred_best_threshold, matching_test_split$prognosis) # Create confusion matrix table
 dimnames(pred_table) <- list( # Assign dimension names to match factor levels

### 2024-10-18 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Added visualisation before and after imputation

@@
 
+# Load the function for PCoA
+source("~/University/MSB/MSB1015/funPCoA.R") # Change the directory as needed
+
+# Load the confusion function
+source("~/University/MSB/MSB1015/confusionPlot.R")
+
 # Load the data
 data <- read_csv("METABRIC_RNA_Mutation.csv")
+
 ```
@@
 any(duplicated(data))
+
+# Call the function for the unsupervised random forest PCA
+PCoA_plots_rawdata <- PCoA_proximity_plot(data, "tumor_stage")
+grid.arrange(grobs = PCoA_plots_rawdata, ncol = 2)
 
@@
 
-# Select only the transcription data
-pca_gx <- data_filled %>%
-  select(brca1:ugt2b7)
-
-# Perform PCA to check the columns with the highest amount of missing values
-pca <- prcomp(pca_gx, center = TRUE, scale. = TRUE)
-
-# Convert PCA results to data frame
-pca_data <- as.data.frame(pca$x[, 1:3])
-pca_data$tumor_stage <- as.factor(data_filled$tumor_stage)
-
-# Create pairwise plots using ggpairs
-ggpairs(pca_data,
-  mapping = aes(color = tumor_stage),
-  upper = list(continuous = wrap("points")),
-  lower = list(continuous = wrap("points")),
-  diag = list(continuous = wrap("barDiag"))
-) +
-  theme_minimal()
 ```
@@
 imputed_data <- read_csv("imputed_data.csv")
+
+# Call the function for the unsupervised random forest PCoA 
+PCoA_plots_rawdata <- PCoA_proximity_plot(data, "tumor_stage", n_trees = 10)
+grid.arrange(grobs = PCoA_plots_rawdata, ncol = 2)
 
@@
 
-# Load the function for PCA
-source("~/University/MSB/MSB1015/funPCA.R") # Change the directory as needed
-
-# Call the function for the unsupervised random forest PCA
-pca_plots_categorical <- pca_proximity_plot(categorical_data, "prognosis")
-grid.arrange(grobs = pca_plots_categorical, ncol = 2)
+# Call the function for the unsupervised random forest PCoA
+PCoA_plots_categorical <- PCoA_proximity_plot(categorical_data, "prognosis")
+grid.arrange(grobs = PCoA_plots_categorical, ncol = 2)
 
@@
 
-# Call the function for the unsupervised random forest PCA (will take a few minutes)
-pca_plots_genetic <- pca_proximity_plot(genetic_data, "prognosis")
-grid.arrange(grobs = pca_plots_genetic, ncol = 2)
+# Call the function for the unsupervised random forest PCoA (will take a few minutes)
+PCoA_plots_genetic <- PCoA_proximity_plot(genetic_data, "prognosis")
+grid.arrange(grobs = PCoA_plots_genetic, ncol = 2)
 
@@
 
-# Call the function for the unsupervised random forest PCA (will take a few minutes)
-pca_plots_complete <- pca_proximity_plot(complete_data, "prognosis")
-grid.arrange(grobs = pca_plots_complete, ncol = 2)
+# Call the function for the unsupervised random forest PCoA (will take a few minutes)
+PCoA_plots_complete <- PCoA_proximity_plot(complete_data, "prognosis")
+grid.arrange(grobs = PCoA_plots_complete, ncol = 2)
 ```
@@
 # Call the plotting function for the confusion matrix
-source("~/University/MSB/MSB1015/confusionPlot.R") # Load external plot function (update directory if needed)
 confusionPlot(train_confusion, title = "Confusion Matrix (Training Set)") # Plot confusion matrix

### 2024-10-18 Code/MSB1015-Project.Rmd
Update MSB1015-Project.Rmd
Removed the number of trees from the PCoA function call

@@
 # Call the function for the unsupervised random forest PCoA 
-PCoA_plots_rawdata <- PCoA_proximity_plot(data, "tumor_stage", n_trees = 10)
+PCoA_plots_rawdata <- PCoA_proximity_plot(data, "tumor_stage")
 grid.arrange(grobs = PCoA_plots_rawdata, ncol = 2)

### 2024-10-18 Code/funPCA.R
Update funPCA.R
Added some code to make the variable a factor for better visualisation and changed the naming system

@@ -1,53 +1,59 @@
-pca_proximity_plot <- function(data, outcome_var, n_trees = 2000, mtry_value = 2, pc_pairs = 1:4) {
+library(caret)
+library(ggplot2)
+library(RColorBrewer)
+
+PCoA_proximity_plot <- function(data, outcome_var, n_trees = 1000, mtry_value = 2, pc_pairs = 1:4) {
+  # Remove rows with NA values from the data
+  data_complete <- data[complete.cases(data), ]
+  
   # Ensure outcome_var is treated as a column name
-  outcome_data <- data[[outcome_var]]
+  outcome_data <- data_complete[[outcome_var]]
   
   # Build Random Forest model with proximity calculation
-  rf_model <- randomForest(x = data[, !names(data) %in% outcome_var], 
+  rf_model <- randomForest(x = data_complete[, !names(data_complete) %in% outcome_var], 
                            mtry = mtry_value, 
                            ntree = n_trees, 
                            proximity = TRUE)
   
-  # Extract proximity matrix and assign row/column names based on the outcome variable
+  # Extract proximity matrix
   proximity_mat <- rf_model$proximity
-  rownames(proximity_mat) <- outcome_data
-  colnames(proximity_mat) <- outcome_data
   
-  # Function to perform PCA and plot for specified PCs
-  pca_plot <- function(data, pc_x, pc_y) {
-    pca_res <- prcomp(data, center = TRUE, scale. = TRUE)
+  # Function to perform PCoA and plot for specified PCs
+  PCoA_plot <- function(data, pc_x, pc_y) {
+    PCoA_res <- prcomp(data, center = TRUE, scale. = TRUE)
     
     # Create a data frame for the selected principal components
-    pca_df <- data.frame(
-      PCx = pca_res$x[, pc_x],
-      PCy = pca_res$x[, pc_y],
-      Outcome = outcome_data
+    PCoA_df <- data.frame(
+      PCx = PCoA_res$x[, pc_x],
+      PCy = PCoA_res$x[, pc_y],
+      Outcome = as.factor(outcome_data)
     )
     
     # Calculate the explained variance
-    explained_var <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100
[…]

### 2024-10-18 Code/funPCA.R
Update funPCA.R

@@ -1,4 +1,13 @@
-library(caret)
+# List of additional required packages
+additional_packages <- c("ggplot2", "RColorBrewer")
+
+# Check and install missing packages
+new_additional_packages <- additional_packages[!(additional_packages %in% installed.packages()[,"Package"])]
+if (length(new_additional_packages)) {
+  install.packages(new_additional_packages)
+}
+
+# Load the libraries
 library(ggplot2)
 library(RColorBrewer)
 
